---
title: "Municipal Finance: Revenue share over time"
format:
  html:
    code-fold: true
    code-summary: "Code"
---

```{r, global_options, include=FALSE}
knitr::opts_chunk$set(tidy.opts = list(width.cutoff = 60), tidy = TRUE)
```

I recently needed to make this graph for something, so I'm putting it on here as well. Unfortunately, it isn't very replicable, because I manually edited which statistics canada data I wanted from the website instead of downloading it and then modifying it here or using a package to grab the code directly. 

Maybe one day I'll fix that. I want to do a more extensive summary look at Canadian municipal finances at some point, but this is it for now.

```{r, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(lubridate)
library(tidyr)

revenue_df <- read.csv("C:/Users/willi/Downloads/3610047701_databaseLoadingData.csv")

main_data <- revenue_df %>%
  filter(Levels.of.government %in% c("General governments", "Federal general government", "Provincial and territorial general governments", "Local general governments")) %>%
  select(REF_DATE, Levels.of.government, Estimates, VALUE)

revenue_data <- main_data %>%
  filter(Estimates == "General governments revenue")

general_revenue_data <- revenue_data %>%
  filter(Levels.of.government == "General governments")

local_revenue_data <- revenue_data %>%
  filter(Levels.of.government == "Local general governments")

provincial_revenue_data <- revenue_data %>%
  filter(Levels.of.government == "Provincial and territorial general governments")

federal_revenue_data <- revenue_data %>%
  filter(Levels.of.government == "Federal general government")



wide_revenue_data <- revenue_data %>%
  pivot_wider(
    names_from = Levels.of.government, 
    values_from = VALUE,
    names_prefix = "Revenue_"
  )


percentage_revenue_data <- wide_revenue_data %>%
  mutate(across(starts_with("Revenue_"), 
                ~ (. / `Revenue_General governments`) * 100, 
                .names = "Percent_{.col}")) %>%
  select(REF_DATE, starts_with("Percent_"))

long_revenue_data <- percentage_revenue_data %>%
  pivot_longer(
    cols = starts_with("Percent_"),
    names_to = "Government_Level",
    values_to = "Percentage"
  ) %>%
  mutate(Government_Level = gsub("Percent_Revenue_", "", Government_Level))


percentage_revenue_data <- percentage_revenue_data %>%
  mutate(REF_DATE = as.Date(paste0(REF_DATE, "-01"), format = "%Y-%m-%d")) # Add a day and convert to Date

ggplot(data = percentage_revenue_data, aes(x = REF_DATE, y = `Percent_Revenue_Local general governments`)) +
  geom_point() +
  scale_x_date(
    date_breaks = "5 years", # Display a tick every 5 years
    date_labels = "%Y"       # Display only the year on the x-axis
  ) +
  labs(
    title = "Local Government Revenue as a Percentage of General Government Revenue",
    x = "Year",
    y = "Percentage of Revenue"
  ) +
  theme_minimal()

```